The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-03-13 paper/iphas-dr2.tex

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage{amssymb,amsmath}
\usepackage[pdftex]{graphicx}
\usepackage{longtable} 
%\usepackage{lscape}
\usepackage{float}
\usepackage{booktabs}
%\usepackage{aas_macros}

@@

Like any other photometric survey,
the majority of the objects in our catalogue
are faint sources observed near the detection limits:
55\% of the entries in the catalogue
are fainter than $r > 20$
and 18\% are even fainter than $r > 21$.
The measurements of such faint objects
are naturally prone to large
random and systematic uncertainties,
for example, an inaccurately subtracted background
will introduce a proportionally larger systematic error
in a faint object.
Most scientific applications will require a set of
quality criteria to be enforced for the purpose
of removing faint and low-quality objects.

@@

\subsection{Photometric uncertainties}

Figure~\ref{fig:uncertainties} shows the mean photometric
uncertainties as a function of magnitude for each band.
We find that the uncertainties typically
reach 0.1 mag near 20.5 in $r$ 
and 19.5 in $i$/\ha when the default 2.3\arcsec\ aperture is used.
At this point we note that the average colour
of objects in the survey is
$1.06\pm0.12$ for ($r$-$i$) and $0.44\pm0.03$ for ($r$-\ha).
The better depth of $r$ is hence compensated
by the fact that stars tend to have 
brighter magnitudes in $i$ and \ha.

@@

\begin{figure*}
    \vspace{1cm}
    \includegraphics[width=0.5\textwidth]{./plots/pairmag/pairmag.pdf} 
    \caption{Photometric repeatability illustrated by plotting
             the mean residuals between all the primary and secondary detections
             in the catalogue as a function of magnitude.
             The best photometric repeatability is reached at $r=14$
             with $\sigma_{r-r2}~=~0.041$~mag.
             Note that bright stars at $r<13$ and $i<12$ 
             show increasing uncertainties due to saturation effects.}
    \label{fig:pairmag}
    \vspace{1cm}
    \includegraphics[width=0.5\textwidth]{./plots/pairmag/pairmag-reliable.pdf} 
    \caption{Same as Figure~\ref{fig:pairmag},
    except that only the subset of sources
    flagged as \emph{veryReliable} are now included.
    We find that applying the quality criteria
    has improved the photometric repeatability significantly.
    The best repeatability is again reached at $r=14$
    but has reduced to $\sigma=0.028$ mag.
    The quality criteria have also been successful
    at removing objects with large systematics at the bright
    and faint ends.}
    \label{fig:pairmag_reliable}
\end{figure*}

@@

\subsection{Source densities and blending problems}
\label{sec:densities}

@@

This is demonstrated in Figures \ref{fig:l180}, \ref{fig:l45} \& \ref{fig:l30},
where we present three sets of IPHAS colour/magnitude diagrams
towards three distinct sightlines
located at Galactic longitudes $180^\circ$, $45^\circ$ and $30^\circ$,
respectively.
Each figure contains all the sources which are flagged as \emph{veryReliable}
and are located in a region of one square degree 
centred on the coordinates indicated in the diagram
(i.e. within a radius of $0.564^\circ$ from the indicated sightline).
For clarity, we have imposed the additional criterion
for photometric uncertainties to be smaller than 0.05 mag in each band
(corresponding to a magnitude limit near $\sim$19th magnitude, effectively).

Each of the diagrams reveals a well-defined locus,
which demonstrates the health of the catalogue and the global calibration
for investigating stellar populations across wide areas.
We have annotated the colour-colour diagrams
by showing the position of the unreddened main sequence (thin solid line),
the unreddened giant branch (thick solid line)
and the reddening track for an A0V-type star (dashed line)
-- all three are based on the \cite{Pickles1998} library of empirical spectra
tabulated for IPHAS by \cite{Drew2005}.
In the colour-magnitude diagrams we only show the reddening vector
together with the unreddened 1~Gyr isochrone due to \cite{Bressan2012},
which are made available for the IPHAS filter system through a
popular on line tool hosted by the Observatory of Padova (http://stev.oapd.inaf.it/cmd).
The isochrone and reddening vector has been placed
at an arbitrary distance of 2~kpc.

Each of the sightlines reveals a stellar population
with distinct characteristics.
Firstly, towards the Galactic anti-centre at $l=180^\circ$ (Fig.~\ref{fig:l180}) 
we find a population dominated by lowly-reddened main sequence stars.
This is consistent with the estimated total sightline extinction of $E(B-V)=0.49$
given by \cite{Schlegel1998}, following the correction of \cite{Schlafly2011}.
Looking in more detail we can see that the stellar locus is narrower for M-type dwarfs than for earlier types;
we do not observe M dwarfs experiencing the strongest reddening possible for this sightline.
This implies that extinction is still increasing at distances beyond $\sim 2$~kpc,
where M dwarfs are too faint to be contained in the IPHAS catalogue.
It is also clear that there are no unreddened stars earlier than $\sim$K0 visible,
such stars would be saturated if within a few hundred parsecs.
This therefore suggests that there is a measurable increase in extinction locally.
We also note an relative absence of late type giants which,
due to the relative brevity of the corresponding phase of stellar evolution,
would only account for a small proportion of a volume limited sample.

@@

Finally, in one of our earliest sightlines at $l=30^{\circ}$
we find a very high number of extremely reddened giants
in addition to an unreddened population
of foreground dwarfs.
In contrast to the sightline at $l=45^{\circ}$,
there is no clear group of giant stars visible in the colour-magnitude diagram of Fig.~\ref{fig:l30},
though the red clump stars are manifested as a slight over density which sits roughly 0.4~mags redder than the A0V reddening track.
At $(l,b)=(45^{\circ}, 2^{\circ})$ the giant stars observed exhibit a relatively narrow range of reddenings as the lie beyond most or essentially all of the Galactic dust.
At $(l,b)=(30^{\circ}, 0^{\circ})$ this is not the case:
even at the substantial distances at which we can observe reddened giant stars extinction is continuing to rise and we still find ourselves well within the Galactic dust disc.
It is also apparent that the $(r-i)$ width of both the M dwarfs and early A dwarfs is greater than that in Fig.~\ref{fig:l45}.
This is indicative of a steeper rise in reddening,
both within several hundred parsecs (M dwarfs)
and within a few kpc (early A dwarfs).

@@

\begin{figure*}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-180-3.pdf} 
        \includegraphics[width=0.5\textwidth]{./plots/cmd-180-3.pdf}
    \end{minipage}
    \caption{Colour-colour and colour-magnitude diagram (left and right panel)
    showing sources flagged as \emph{veryReliable}
    located in an area of one square degree
    centred near the Galactic anti-centre 
    at $(l,b)=(180^\circ,+3^\circ)$.
    The colour-colour diagram shows the
    position of the main sequence (thin solid line),
    giant stars (thick solid line)
    and the reddening track for an A0V-type star (dashed line)
    based on the \citet{Pickles1998} library of empirical spectra.
    The colour-magnitude only show the reddening vector
    along with the unreddened 1~Gyr isochrone due to \citet{Bressan2012},
    which has been placed at an arbitrary distance of 2~kpc for reference.
    This is one of the least reddened sightlines
    in the survey %(Av XX, Schlegel et al. XXXX)
    and hence the observed stellar population appears to be dominated 
    by lowly reddened main sequence stars.}
    \label{fig:l180}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-45-2.pdf}
        \includegraphics[width=0.5\textwidth]{./plots/cmd-45-2.pdf}
    \end{minipage}
    \caption{Same as above for $(l,b)=(45^\circ,+2^\circ)$,
    which is one of the highest-density sightlines in the survey,
    revealing two groups of stars in colour-magnitude space.}
    \label{fig:l45}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-30-0.pdf}
        \includegraphics[width=0.5\textwidth]{./plots/cmd-30-0.pdf} 
    \end{minipage}
    \caption{Same as above for $(l,b)=(30^\circ,0^\circ)$.
    This is one of the most reddened sightlines in the survey.
    %(XXX, Schlegel et al XXXX).
    }
    \label{fig:l30}
\end{figure*}

@@

Preparation of the catalogue was eased greatly
by a number of software packages,
including the Python modules
AstroPy \citep{Astropy},
APLpy, NumPy and SciPy,
the PostgreSQL database software,
the TOPCAT and STILTS packages \citep{Taylor2005,Taylor2006},
and the Montage software maintained by NASA/IPAC.
We also made use of the SIMBAD, Vizier and Aladin \citep{Aladin} tools operated at CDS, Strasbourg, France.

@@

\bibliographystyle{mn2e}
\bibliography{dr2}

>>>>

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage{amssymb,amsmath}
\usepackage[pdftex]{graphicx}
\usepackage{longtable} 
%\usepackage{lscape}
\usepackage{float}
\usepackage{booktabs}
\usepackage{aas_macros}

@@

Like any other photometric survey,
the majority of the objects in our catalogue
are faint sources observed near the detection limits:
55\% of the entries in the catalogue
are fainter than $r > 20$
and 18\% are even fainter than $r > 21$.
The measurements of such faint objects
are naturally prone to large
random and systematic uncertainties,
for example, an inaccurately subtracted background
will introduce a proportionally larger systematic error
in a faint object.
Most scientific applications will require a set of
quality criteria to be enforced for the purpose
of removing such faint and low-quality objects.

@@

\subsection{Photometric uncertainties and repeatability}

Figure~\ref{fig:uncertainties} shows the mean photometric
uncertainty as a function of magnitude for each band.
We find that the uncertainty typically
reaches 0.1 mag near 20.5 in $r$ 
and 19.5 in $i$/\ha when the default 2.3\arcsec\ aperture is used.
At this point we note that the average colour
across all sources in the catalogue is
$1.06\pm0.12$ for ($r$-$i$) and $0.44\pm0.03$ for ($r$-\ha).
The better depth of $r$ is hence compensated
by the fact that stars tend to have 
brighter magnitudes in both $i$ and \ha.

@@

\begin{figure*}
\begin{minipage}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{./plots/pairmag/pairmag.pdf}
\label{fig:pairmag}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{./plots/pairmag/pairmag-reliable.pdf}
\label{fig:pairmag_reliable}
\end{minipage}
\caption{Photometric repeatability as a function of magnitude
         for both all sources in the catalogue (left panel)
         and for the \emph{veryReliable} sources alone (right panel).
         The values shown are the mean residuals
         between the primary and the secondary detections,
         while the errorbars show the standard deviations.
         Left panel: 
         the best photometric repeatability is reached at $r=14$
         with $\sigma_{r-r2}~=~0.041$~mag.
         Bright stars at $r<13$ and $i<12$ 
         show increasing uncertainties due to saturation effects.
         Right panel: we find that applying the quality criteria
         has improved the photometric repeatability significantly.
         The best repeatability is again reached at $r=14$
         but has reduced to $\sigma=0.028$ mag.
         The quality criteria have also been successful
         at removing objects with large systematics at the bright
         and faint ends.
    }
\end{figure*}

@@

\subsection{Source densities and blending}
\label{sec:densities}

@@

This is demonstrated 
in Figures \ref{fig:l180}, \ref{fig:l45} \& \ref{fig:l30},
where we present three sets of IPHAS colour/magnitude diagrams
towards three distinct sightlines
located at Galactic longitudes 
$180^\circ$, $45^\circ$ and $30^\circ$, respectively.
Each figure contains all the sources
which are flagged as \emph{veryReliable}
and are located in a region of one square degree 
centred on the coordinates indicated in the diagram
(i.e. within a radius of $0.564^\circ$ from the indicated sightline).
For clarity, we have imposed the additional criterion
that the photometric uncertainties
must be smaller than 0.05 mag in each band
(corresponding to a magnitude limit near $\sim$19th magnitude).

Each of the diagrams reveals a well-defined locus,
which demonstrates the health of the catalogue and the global calibration
for investigating stellar populations across wide areas.
We have annotated the colour-colour diagrams
by showing the position 
of the unreddened main sequence (thin solid line),
the unreddened giant branch (thick solid line),
and the reddening track for an A0V-type star (dashed line)
-- all three are based on the \cite{Pickles1998} library 
of empirical spectra
transformed into the Vega-based IPHAS system by \cite{Drew2005}.
In the colour-magnitude diagrams we only show the reddening vector
together with the unreddened 1~Gyr isochrone due to \cite{Bressan2012},
which is made available for the IPHAS system through a
popular on line tool hosted at the Observatory of Padova
(http://stev.oapd.inaf.it/cmd).
The isochrone and reddening vector have been placed
at an arbitrary distance of 2~kpc.

Each of the sightlines reveal a stellar population
with distinct characteristics.
Firstly, towards the Galactic anti-centre 
at $l=180^\circ$ (Fig.~\ref{fig:l180})
we find a population dominated by lowly-reddened main sequence stars.
This is consistent with the estimated total sightline extinction 
of $E(B-V)=0.49$ given by \cite{Schlegel1998},
following the correction of \cite{Schlafly2011}.
Looking in more detail we can see
that the stellar locus is narrower for M-type dwarfs
than for earlier types;
we do not observe M dwarfs experiencing 
the strongest reddening possible for this sightline.
This implies that extinction is still increasing 
at distances beyond $\sim$1-2~kpc,
where M dwarfs are too faint to be contained in the IPHAS catalogue.
It is also clear that there are no unreddened stars
earlier than $\sim$K0 visible,
such stars would be saturated if within a few hundred parsecs.
This therefore suggests 
that there is a measurable increase in extinction locally.
We also note a relative absence of late type giants which,
due to the relative brevity 
of the corresponding phase of stellar evolution,
would only account for a small proportion of a volume limited sample.

@@

Finally, in one of our earliest sightlines at $l=30^{\circ}$
we find a very high number of extremely reddened giants
in addition to an unreddened population
of foreground dwarfs.
In contrast to the sightline at $l=45^{\circ}$,
there is no clear group of giant stars visible
in the colour-magnitude diagram of Fig.~\ref{fig:l30},
though the red clump stars are manifested as a slight over density
which sits roughly 0.4~mags redder than the A0V reddening track.
At $(l,b)=(45^{\circ}, 2^{\circ})$ the giant stars observed
exhibit a relatively narrow range of reddenings
as they lie beyond most or essentially all of the Galactic dust.
At $(l,b)=(30^{\circ}, 0^{\circ})$ this is not the case:
even at the substantial distances
at which we can observe reddened giant stars,
extinction is continuing to rise
and we still find ourselves well within the Galactic dust disc.
It is also apparent that the $(r-i)$ width of both the M dwarfs
and early A dwarfs is greater than that in Fig.~\ref{fig:l45}.
This is indicative of a steeper rise in reddening,
both within several hundred parsecs (M dwarfs)
and within a few kpc (early A dwarfs).

@@

\begin{figure*}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-180-3.pdf} 
        \includegraphics[width=0.5\textwidth]{./plots/cmd-180-3.pdf}
    \end{minipage}
    \caption{Colour-colour and colour-magnitude diagrams
    (left and right panel)
    showing sources flagged as \emph{veryReliable}
    located in an area of one square degree
    centred near the Galactic anti-centre 
    at $(l,b)=(180^\circ,+3^\circ)$.
    The diagrams are plotted as 2D-histograms
    which show the density of objects
    in bins of $\sim0.01\times0.01$~mag;
    bins containing 1-10 objects are coloured red,
    bins with 10-20 objects are orange,
    and bins with 20-25 objects are yellow.
    The left panel is annotated with
    the position of the main sequence (thin solid line),
    giant stars (thick solid line)
    and the reddening track for an A0V-type star (dashed line),
    which are based on the \citet{Pickles1998} library
    of observed spectra.
    The right panel only shows the reddening vector
    along with the unreddened 1~Gyr isochrone due to \citet{Bressan2012},
    which has been placed at an arbitrary distance of 2~kpc.
    This is one of the least reddened sightlines
    in the survey
    and hence the observed stellar population appears to be dominated 
    by lowly reddened main sequence stars (see text).}
    \label{fig:l180}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-45-2.pdf}
        \includegraphics[width=0.5\textwidth]{./plots/cmd-45-2.pdf}
    \end{minipage}
    \caption{Same as above for $(l,b)=(45^\circ,+2^\circ)$,
    which is one of the highest-density sightlines in the survey,
    revealing two groups of stars in colour-magnitude space.}
    \label{fig:l45}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-30-0.pdf}
        \includegraphics[width=0.5\textwidth]{./plots/cmd-30-0.pdf} 
    \end{minipage}
    \caption{Same as above for $(l,b)=(30^\circ,0^\circ)$,
    showing one of the most reddened sightlines in the survey.
    }
    \label{fig:l30}
\end{figure*}

@@

Preparation of the catalogue was eased greatly
by a number of software packages,
including the Python modules
AstroPy \citep{Astropy},
APLpy, NumPy and SciPy,
the PostgreSQL database software,
the TOPCAT and STILTS packages \citep{Taylor2005,Taylor2006},
and the Montage software maintained by NASA/IPAC.
We also made use of the SIMBAD, Vizier and Aladin \citep{Aladin}
operated at CDS, Strasbourg, France.

@@

\bibliographystyle{mn2e}
\bibliography{iphas-dr2}
